---
# TODO Update the title once we have the findings
title: "Impact of Weather Events to Health and Economy in United States"
author: "A. Johan"
output: html_document
---

## Synopsis

-- TODO --

## Introduction 

The final project for the course Reproducible Research is the study of the 
impact of weather events on population health and economic consequences. 
To do so, we will explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The explanation of the data can be found at the following links:

- [National Weather Service Storm Data Documentation][1]
- [National Climatic Data Center Storm Events FAQ][2]

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

## Data Processing

For this analysis we will be using the following libraries:

- dplyr

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load the libraries 
library(dplyr)
```


### Downloading and reading the data

Source data is provided by the course website and can be downloaded here:
[Storm Data][3]

[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

The source data is compressed in bzip2 format. It is a CSV file. 
Fortunately R has the capability to directly read CSV files compressed in 
bzip2 format.

```{r, cache=TRUE}
# download source data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dest <- "stormdata.csv.bz2"
rdsfile <- "stormdata.rds"

# Download the file and save it to dest if it not present in the folder
if (!file.exists(dest)){
  download.file(url = url, destfile = dest, quiet = TRUE)
}

# TODO uncomment below prior to submission
# data <- as_tibble(read.csv(dest))

# TODO remove this before submissions
if (!file.exists(rdsfile)) {
  saveRDS(read.csv(dest), file = rdsfile)
}
if(length(ls(pattern = "data")) == 0) {
data <- as_tibble(readRDS(rdsfile))
}
```

Initial inspection of the data:

- There are `r nrow(data)` observations with `r ncol(data)` variables
- Some of the event labels are in uppercase and lowercase
text which causes R to treat them differently. Some EVTYPE labels also include
white spaces. To simplify the analysis, leading and trailing white spaces
will be removed and all of the event type names will be converted to uppercase

```{r, cache=TRUE}
# Trim white spaces and convert all EVTYPE to upper case
data$EVTYPE <- toupper(trimws(data$EVTYPE))
```

### Data Cleaning and Transformation

To see effect of event to fatalities and injuries in the population we can 
summarize the data by the total  fatalities, total injuries, frequency of events.
We can then combine total fatalities and total injuries into a single variable called
total_casualty.

```{r, cache=TRUE}
health_summary <- data %>% 
  group_by(EVTYPE) %>% 
  summarise(total_fatalities = sum(FATALITIES), 
            total_injuries = sum(INJURIES), .groups = "keep")

ev_freq <- data %>% group_by(EVTYPE) %>% tally()

# Merge the count and then calculate the total casualty 
health_summary <- merge(health_summary, ev_freq, by = "EVTYPE") %>%
  mutate(total_casualty = total_fatalities + total_injuries)
```

To see the effect of the events on the economic factors we can do a summary by 
the variable `PROPDMG` and `CROPDMG`. Care should be taken however since
there is another variable called `PROPDMGEXP` and `CROPDMGEXP` which seems to
imply a scalar multiplication to the amount.

We will refer to the analysis done on [this following website][4] visited 
on July 6, 2020 to translate the exponential values.

[4]: https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html

To summarize, the exponential values are:

|Label|Multiplier|
|-----|----------|
|H,h | 100|
|K,k | 1,000 |
|M,m | 1,000,000|
|B,b | 1,000,000,000|
|Numeric 0..8 | 10 |
|+ | 1|
|- | 0|
|? | 0|
|Blank/empty character | 0

We will normalize the crop and property damage value by creating a new variable 
`PROPAMT` and `CROPAMT` which will contain the respective `xDMG` variables 
multiplied by the scalar in `xDMGEXP`.

An important point to note is that we should substitute the `numeric 0..8` values
first to avoid confusion when substituting for the `+` symbol. 

```{r}
# Since the list of values to substitute is not a lot we can do this 
# individually. to simplify convert all to uppercase
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$CROPDMGEXP <- toupper(data$CROPDMGEXP)

# NOTE: we must replace with character which we will cast to numeric after all 
# substitution is done

# replace the numeric values with 10
data[data$PROPDMGEXP == as.character(c(1:8)), "PROPDMGEXP"] <- "10"

# replace the 0 values 
data[data$PROPDMGEXP == "", "PROPDMGEXP"] <- "0"
data[grepl("-|\\?", x$PROPDMGEXP), "PROPDMGEXP"] <- "0"

```


## Results

-- TODO --
