---
title: "Tornado and Floods as Major Impactors to Health and Economy in United States"
author: "A. Johan"
output: html_document
---

## Synopsis

Using the data provided by the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, we can summarize the data to identify
the events that adversely impact the population in terms of fatality, injuries
as well as events that cause the greatest damage to property and crops in the 
United States.

The findings of this analysis is that **Tornado** causes the greatest
loss of life and injury. 

**Flood** causes the highest economic loss overall, crops
are more affected by **Drought**.

## Introduction 

The final project for the course Reproducible Research is the study of the 
impact of weather events on population health and economic consequences. 
To do so, we will explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The questions we are trying to answer are:

- Across the United States, which types of events are most harmful with 
respect to population health?
- Across the United States, which types of events have the greatest economic consequences?

The explanation of the data can be found at the following links:

- [National Weather Service Storm Data Documentation][1]
- [National Climatic Data Center Storm Events FAQ][2]

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf

## Data Processing

For this analysis we will be using the following libraries:

- dplyr
- ggplot2

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Load the libraries 
library(dplyr)
library(ggplot2)
```


### Downloading and reading the data

Source data is provided by the course website and can be downloaded here:
[Storm Data][3]

[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

The source data is compressed in bzip2 format. It is a CSV file. 
Fortunately R has the capability to directly read CSV files compressed in 
bzip2 format.

```{r}
# download source data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dest <- "stormdata.csv.bz2"
rdsfile <- "stormdata.rds"

# Download the file and save it to dest if it not present in the folder
if (!file.exists(dest)){
  download.file(url = url, destfile = dest, quiet = TRUE)
}

data <- as_tibble(read.csv(dest))
```

### Data Cleaning and Transformation

Initial inspection of the data:

- There are `r nrow(data)` observations with `r ncol(data)` variables
- Some of the event labels are in uppercase and lowercase
text which causes R to treat them differently. Some EVTYPE labels also include
white spaces. To simplify the analysis, leading and trailing white spaces
will be removed and all of the event type names will be converted to uppercase

```{r}
# Trim white spaces and convert all EVTYPE to upper case
data$EVTYPE <- toupper(trimws(data$EVTYPE))
```

#### Fatalities & Injuries
To see effect of event to fatalities and injuries in the population we can 
summarize the data by the total  fatalities, total injuries, frequency of events.
We can then combine total fatalities and total injuries into a single variable called
total_casualty.

```{r}
health_summary <- data %>% 
  group_by(EVTYPE) %>% 
  summarise(TF = sum(FATALITIES), 
            TI = sum(INJURIES), .groups = "keep")  %>%
  mutate(TOTAL = TF + TI)

ev_freq <- data %>% group_by(EVTYPE) %>% tally(name = "COUNT")

# Merge the count and then calculate the total casualty 
health_summary <- merge(health_summary, ev_freq, by = "EVTYPE")
```

#### Property and Crop Damage

To see the effect of the events on the economic factors we can do a summary by 
the variable `PROPDMG` and `CROPDMG`. Care should be taken however since
there is another variable called `PROPDMGEXP` and `CROPDMGEXP` which seems to
imply a scalar multiplication to the amount.

We will refer to the analysis done on [this following website][4] visited 
on July 6, 2020 to translate the exponential values.

[4]: https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html

To summarize, the exponential values are:

|Label|Multiplier|
|-----|----------|
|H,h | 100|
|K,k | 1,000 |
|M,m | 1,000,000|
|B,b | 1,000,000,000|
|Numeric 0..8 | 10 |
|+ | 1|
|- | 0|
|? | 0|
|Blank/empty character | 0

We will normalize the crop and property damage value by creating a new variable 
`PROPAMT` and `CROPAMT` which will contain the respective `xDMG` variables 
multiplied by the scalar in `xDMGEXP`.

An important point to note is that we should substitute the `numeric 0..8` values
first to avoid confusion when substituting for the `+` symbol. 

```{r}
# Since the list of values to substitute is not a lot we can do this 
# individually. to simplify convert all to uppercase
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$CROPDMGEXP <- toupper(data$CROPDMGEXP)

# NOTE: we must replace with character which we will cast to numeric after all 
# substitution is done

# replace the numeric values with 10
data[data$PROPDMGEXP %in% as.character(c(1:8)), "PROPDMGEXP"] <- "10"
data[data$CROPDMGEXP %in% as.character(c(1:8)), "CROPDMGEXP"] <- "10"

# replace the 0 values 
data[data$PROPDMGEXP == "", "PROPDMGEXP"] <- "0"
data[data$CROPDMGEXP == "", "CROPDMGEXP"] <- "0"

data[grepl("-|\\?", data$PROPDMGEXP), "PROPDMGEXP"] <- "0"
data[grepl("-|\\?", data$CROPDMGEXP), "CROPDMGEXP"] <- "0"

# replace +
data[data$PROPDMGEXP == "+", "PROPDMGEXP"] <- "1"
data[data$CROPDMGEXP == "+", "CROPDMGEXP"] <- "1"

# replace H
data[data$PROPDMGEXP == "H", "PROPDMGEXP"] <- "100"
data[data$CROPDMGEXP == "H", "CROPDMGEXP"] <- "100"

# replace K
data[data$PROPDMGEXP == "K", "PROPDMGEXP"] <- "1000"
data[data$CROPDMGEXP == "K", "CROPDMGEXP"] <- "1000"

# replace M
data[data$PROPDMGEXP == "M", "PROPDMGEXP"] <- "1000000"
data[data$CROPDMGEXP == "M", "CROPDMGEXP"] <- "1000000"

# replace B
data[data$PROPDMGEXP == "B", "PROPDMGEXP"] <- "1000000000"
data[data$CROPDMGEXP == "B", "CROPDMGEXP"] <- "1000000000"

# create the actual amount in USD for property and crop dmg. we will also
# create a new variable which is the sum of both
data <- data %>% 
  mutate(PROPAMT = as.numeric(PROPDMGEXP) * PROPDMG) %>%
  mutate(CROPAMT = as.numeric(CROPDMGEXP) * CROPDMG) %>%
  mutate(TOTALAMT = CROPAMT + PROPAMT)
```

Once we have normalized the crop and property damage values we can then proceed
to summarize the data

```{r}
econ_summ <- data %>% 
  group_by(EVTYPE) %>% 
  summarise(TP = sum(PROPAMT), 
            TC = sum(CROPAMT), 
            TOTAL = sum(TOTALAMT), .groups = "keep")

econ_summ <- merge(econ_summ, ev_freq, by = "EVTYPE")
```

The question of interest is with relation to harmful events and economic impact.
With this, we can then exclude events that do not have any health effect or
economic impact.

For economic impact, we only include those with either a property damage or crop
damage amount

For health impact, we include those either with fatality or injury.

For both we exclude unknown events i.e. those with value `?`

```{r}
health_summary <- health_summary %>% filter(EVTYPE != "?")
hs1 <- health_summary %>% filter(TOTAL > 0) %>% select(EVTYPE, TOTAL, COUNT)
hs2 <- health_summary %>% filter(TF > 0) %>% select(EVTYPE, TF, COUNT)
hs3 <- health_summary %>% filter(TI > 0) %>% select(EVTYPE, TI, COUNT)

econ_summ <- econ_summ %>% filter(EVTYPE != "?")
es1 <- econ_summ %>% filter(TOTAL > 0) %>% select(EVTYPE, TOTAL, COUNT)
es2 <- econ_summ %>% filter(TP > 0) %>% select(EVTYPE, TP, COUNT)
es3 <- econ_summ %>% filter(TC > 0) %>% select(EVTYPE, TC, COUNT)
```

## Results

### Fatility and Injury

Looking at the summmary, the event that causes the highest amount of casualty
(injury AND fatility) is:
```{r}
hs1[which.max(hs1$TOTAL), ]
```
The event that causes the highest amount of fatality:
```{r}
hs2[which.max(hs2$TF), ]
```

The event that causes the highest amount of injury:
```{r}
hs3[which.max(hs3$TI), ]
```

Since there are too many events to make a meaningful plot, we will only
plot the top 10 events to see how it compares to each other

```{r}
# get the top 10 total
top10h <- hs1 %>% slice_max(TOTAL, n = 10)

ggplot(top10h, aes(EVTYPE, TOTAL)) + 
    geom_col(fill = "#00abff") + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "Total Casualty for Top 10 Events", 
         x = "", 
         y = "Casualty (Person)")
```

### Property and Crop Damage

The event that causes the highest amount of damage (property AND crop) is:
```{r}
es1[which.max(es1$TOTAL), ]
```

The event that causes the highest amount of property damage:
```{r}
es2[which.max(es2$TP), ]
```

The event that causes the highest amount of crop damage:
```{r}
es3[which.max(es3$TC), ]
```

As before, we will only plot the top 10 events to see how it compares to each other

```{r}
# get the top 10 total
top10 <- es1 %>% slice_max(TOTAL, n = 10)

ggplot(top10, aes(EVTYPE, TOTAL)) + 
    geom_col(fill = "#e74c3c") + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "Economic Impact for Top 10 Events", 
         x = "", 
         y = "Damage (billion USD)") +
  scale_y_continuous(labels = function(x){x/1000000000})
```

## Conclusion

**Tornado** causes the greatest loss of life and injury in the United States.

Overall, **Flood** causes the greatest property damage. However, for crops they
are more likely to be affected by **Drought**.

This analysis, should pave way for next steps to mitigate the effects of flooding
and drouught.
